A star formation event in the leading arm of the Magellanic stream triggered by passage through the outer Galactic disk

### 2018-08-01 ms.tex
Proper motion modeling

% typography
\setlength{\parindent}{1.\baselineskip}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\package}[1]{\textsl{#1}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\pans}{\acronym{PS1}}
\newcommand{\decam}{DECam}
\newcommand{\DR}[1]{\acronym{DR#1}}
\newcommand{\todo}[1]{{\color{red} TODO: #1}}

@@

% astronomy stuffs
\newcommand{\msun}{\textrm{M}_\odot}
\newcommand{\kpc}{\textrm{kpc}}
\newcommand{\kms}{\ensuremath{\textrm{km}~\textrm{s}^{-1}}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\masyr}{\ensuremath{\textrm{mas}~\textrm{yr}^{-1}}}
\newcommand{\feh}{\ensuremath{[\textrm{Fe} / \textrm{H}]}}
%\newcommand{\kms}{\mbox{km s$^{-1}~$}}
\newcommand{\kmse}{\mbox{km s$^{-1}$}}
\newcommand{\vlsr}{$V_{\rm LSR}~$}
\newcommand{\vlsre}{$V_{\rm LSR}$}
\newcommand{\hi}{H{\footnotesize I} }
\newcommand{\hie}{H{\footnotesize I}}
\newcommand{\lms}{$L_{\rm MS}~$}
\newcommand{\lmse}{$L_{\rm MS}$}
\newcommand{\bms}{$B_{\rm MS}~$}
\newcommand{\bmse}{$B_{\rm MS}$}

\newcommand{\clustername}{\textsl{NAME}}
\newcommand{\lmcsmc}{LMC--SMC}
\newcommand{\bprp}{\ensuremath{\textrm{BP} - \textrm{RP}}}

@@

We report the discovery of a young ($\lesssim 100~\textrm{Myr}$), metal-poor ($[\textrm{Fe}/\textrm{H}] \sim -1$) stellar association in the halo of the Milky Way ($(R, z) \sim (24, 16) kpc$).
The association --- \clustername\ --- is likely associated with the leading arm of the gas stream emanating from the Magellanic cloud system, but is located $\approx 60^\circ$ from the Large Magellanic Cloud (LMC) center, on the other side of the Milky Way disk relative to the LMC.
By assuming that the cluster is co-located with HI gas in the stream, we directly measure the distance to the Magellanic stream, $d \approx 25~\textrm{kpc}$.
At this location relative to the LMC, the measured distance is inconsistent with predictions from models of the LMC/SMC interaction and infall into the Milky Way. \todo{that don't account for ram pressure and gas interaction with MW disk}
The estimated age of \clustername\ is consistent with the time of last passage through the Galactic midplane.
We therefore conclude that this star-formation event was triggered by the last disk passage, which occurred at a Galactocentric radius $R \approx XX~\kpc$.

@@

We then query all objects from the \gaia\ \DR{2} catalog within a polygon centered on the nominal position of the group (see above), with a width of $12^\circ$ and a height of $8^\circ$ in the Magellanic stream coordinate system \citep{Nidever:2008} (see Appendix~\ref{sec:queries} for the database query).
We cross-match the resulting catalog with \acronym{GALEX} \todo{citation} using a positional crossmatch requiring source separation $<2~\textrm{arcsec}$, and with \pans\ \todo{citation}\citep{Chambers:XX} using the cross-match provided by the \gaia\ science archive.
\figurename~\ref{fig:cmds} shows the \gaia\ color-magnitude diagram (left) and the \acronym{GALEX} NUV minus \pans\ $g$-band vs. \gaia\ \bprp\ color-color diagram (right) from the cross-matched photometry.
All magnitudes are corrected for dust extinction using the SFD \citep{Schlegel:1998} dust map.
The green polygon in the left panel of \figurename~\ref{fig:cmds} highlights the over-density of blue stars that were identified as comoving from the proper motion search described above.
The red line in right panel of \figurename~\ref{fig:cmds} is an isochrone (from MIST; \citealt{TODO}) for a $50~\textrm{Myr}$, $\feh = -1$ stellar population.

% Notebook: Figure-CMDs
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/gaia-galex-cmd.pdf}
\caption{Color-magnitude and color-color diagrams with \gaia\ ($G$, BP, RP), \acronym{GALEX} (NUV), and \pans\ ($g$) magnitudes.
Shaded pixels (orange) show regions of high density and points (black markers) are plotted when there are fewer than 8 stars per pixel.
Green polygon in left panel highlights the over-density of blue, comoving stars identified by the search described in \sectionname~\ref{sec:discovery}.
Red line in right panel shows the zero-age main sequence of a $50~\textrm{Myr}$, $\feh = -1$ stellar population from the MIST isochrone library \citep{TODO}.
}
\label{fig:cmds}
\end{figure}

@@

\section{Proper motion cleaned}
\label{sec:pmclean}

@@

@@

Query to retrieve \gaia\ data around the blue, comoving group found and discussed in \sectionname~\ref{sec:data}:
\begin{verbatim}
SELECT *
FROM gaiadr2.gaia_source
WHERE (parallax < 0.5 OR parallax IS NULL) AND
    CONTAINS(POINT('ICRS', ra, dec),
             POLYGON('ICRS',
                     182.1954909980203, -35.51467868323814,
                     172.7240896740689, -33.70157733692543,
                     176.3745742549848, -22.277118488646703,
                     184.81885184435987, -24.236022801996057)) = 1 AND
    (pmra > -10) AND (pmra < 10) AND
    (pmdec > -10) AND (pmdec < 10)
\end{verbatim}

\software{
    \package{Astropy} \citep{astropy},
    \package{dustmaps}\footnote{\url{https://github.com/gregreen/dustmaps}},
    \package{gala} \citep{gala},
    \package{IPython} \citep{ipython},
    \package{matplotlib} \citep{mpl},
    \package{numpy} \citep{numpy},
    \package{scipy} \citep{scipy}
}

>>>>

% typography
\setlength{\parindent}{1.\baselineskip}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\package}[1]{\textsl{#1}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\pans}{\acronym{PS1}}
\newcommand{\decam}{DECam}
\newcommand{\DR}[1]{\acronym{DR#1}}
\newcommand{\todo}[1]{{\color{red} TODO: #1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}

@@

% Stats / probability
\newcommand{\given}{\,|\,}
\newcommand{\norm}{\mathcal{N}}

% Maths
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\mat}[1]{\mathbf{#1}}

% Units
\newcommand{\msun}{\textrm{M}_\odot}
\newcommand{\kpc}{\textrm{kpc}}
\newcommand{\kms}{\ensuremath{\textrm{km}~\textrm{s}^{-1}}}
\newcommand{\masyr}{\ensuremath{\textrm{mas}~\textrm{yr}^{-1}}}

% Astronomy
\newcommand{\feh}{\ensuremath{[\textrm{Fe} / \textrm{H}]}}
\newcommand{\vlsr}{$V_{\rm LSR}~$}
\newcommand{\vlsre}{$V_{\rm LSR}$}
\newcommand{\hi}{H{\footnotesize I} }
\newcommand{\hie}{H{\footnotesize I}}
\newcommand{\lms}{$L_{\rm MS}~$}
\newcommand{\lmse}{$L_{\rm MS}$}
\newcommand{\bms}{$B_{\rm MS}~$}
\newcommand{\bmse}{$B_{\rm MS}$}

% Shortcuts for this paper
\newcommand{\clustername}{\textsl{NAME}}
\newcommand{\lmcsmc}{LMC--SMC}
\newcommand{\bprp}{\ensuremath{\textrm{BP} - \textrm{RP}}}
\newcommand{\truepm}{\ensuremath{\tilde{\bs{\mu}}}}

@@

We report the discovery of a young ($\lesssim 100~\textrm{Myr}$), metal-poor ($[\textrm{Fe}/\textrm{H}] \sim -1$) stellar association in the halo of the Milky Way ($(R, z) \sim (24, 16)~\textrm{kpc}$).
The association --- \clustername\ --- is likely associated with the leading arm of the gas stream emanating from the Magellanic cloud system, but is located $\approx 60^\circ$ from the Large Magellanic Cloud (LMC) center, on the other side of the Milky Way disk relative to the LMC.
By assuming that the cluster is co-located with HI gas in the stream, we directly measure the distance to the Magellanic stream, $d \approx 25~\textrm{kpc}$.
At this location relative to the LMC, the measured distance is inconsistent with predictions from models of the LMC/SMC interaction and infall into the Milky Way. \todo{that don't account for ram pressure and gas interaction with MW disk}
The estimated age of \clustername\ is consistent with the time of last passage through the Galactic midplane.
We therefore conclude that this star-formation event was triggered by the last disk passage, which occurred at a Galactocentric radius $R \approx XX~\kpc$.

@@

We then query all objects from the \gaia\ \DR{2} catalog within a rectangle centered on the nominal position of the group (see above), with a width of $12^\circ$ and a height of $12^\circ$ in the equatorial (ICRS) coordinate system (see Appendix~\ref{sec:queries} for the database query).
We cross-match the resulting catalog with \acronym{GALEX} \todo{citation} using a positional crossmatch requiring source separation $<2~\textrm{arcsec}$, and with \pans\ \todo{citation}\citep{Chambers:XX} using the cross-match provided by the \gaia\ science archive.
\figurename~\ref{fig:cmds} shows the \gaia\ color-magnitude diagram (left) and the \acronym{GALEX} NUV minus \pans\ $g$-band vs. \gaia\ \bprp\ color-color diagram (right) from the cross-matched photometry.
All magnitudes are corrected for dust extinction using the SFD \citep{Schlegel:1998} dust map.
The green polygon in the left panel of \figurename~\ref{fig:cmds} highlights the over-density of blue stars that were identified as comoving from the proper motion search described above.
The lines in the right panel of \figurename~\ref{fig:cmds} show isochrones from the \acronym{MIST} (blue; \citealt{TODO}) and \acronym{PARSEC} (red; \citealt{TODO}) isochrone libraries for a 50 Myr stellar population with metallicity $\feh = -0.5$ (dashed) or $\feh = -1.5$ (solid).

% Notebook: Figure-CMDs
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/gaia-galex-cmd.pdf}
\caption{Color-magnitude and color-color diagrams with \gaia\ ($G$, BP, RP), \acronym{GALEX} (NUV), and \pans\ ($g$) magnitudes.
Shaded pixels (orange) show regions of high density and points (black markers) are plotted when there are fewer than 8 stars per pixel.
Green polygon in left panel highlights the over-density of blue, comoving stars identified by the search described in \sectionname~\ref{sec:discovery}.
Red and blue lines in right panel show the zero-age main sequence of a $50~\textrm{Myr}$ stellar population with the metallicity indicated from the \acronym{MIST} \citep{TODO} and \acronym{PARSEC} \citep{TODO} isochrone libraries.
}
\label{fig:cmds}
\end{figure}

@@

\section{Determining kinematic membership}
\label{sec:pmclean}

We identify probable members of the photometrically-identified young cluster using astrometric data from \gaia.
We start by selecting all stars with $\bprp < 0.35$ to remove low-mass and old main sequence star contamination in the region.
\figurename~\ref{fig:pm-members}, left (grey points), shows the sky positions of stars that pass this blue cut in the region around the young cluster.
The larger solid-line (green) circle indicates the region we define as the cluster area, and the two smaller dashed-line (blue) circles indicate control fields that are combined and used for modeling the background distribution of proper motions.
The control fields are designed to, together, have the same total area as the cluster field.
The top-left panel of the middle panels shows the proper motion distribution in the control fields, and top-right panel in the middle shows the proper motion distribution of the cluster field.
The distinct over-density of stars near $(\mu_\alpha, \mu_\delta) \approx (-0.5, 0.5)~\masyr$\footnote{Throughout this article, we use $\mu_\alpha$ to refer to the proper motion value provided by \gaia, which includes the $\cos\delta$ term.} is the identified comoving association of blue stars.

To model membership of the cluster based on the observed proper motions, $\bs{\mu} = (\mu_\alpha, \mu_\delta)$, we first construct a model for the error-deconvolved proper motion distribution in the control fields using ``extreme deconvolution'' \citep[XD;][]{TODO} with 3 Gaussian components.
The bottom-left panel of the middle panels of \figurename~\ref{fig:pm-members} shows the three inferred Gaussian components (blue) from running XD, which takes into account the full error distributions for each proper motion measurement (i.e. including covariances), $\mat{C}_\mu$, provided by \gaia\ \DR{2}.
We fix the parameters of the background model inferred from XD and model the cluster field proper motion distribution using a two-component mixture model with a single, isotropic Gaussian component for the error-deconvolved cluster distribution with mean $\bs{m}$ and isotropic variance $s^2$, and the XD-inferred model for the control field, $p_{\textrm{XD}}$, for the background component.
In detail, taking $f$ to be the fraction of blue stars in this region belonging to the young cluster, and \truepm\ to be the true proper motion for a single star,
\begin{align}
    p(\bs{\mu} \given \truepm, \mat{C}_\mu) &=
        \norm(\bs{\mu} \given \truepm, \mat{C}_\mu)\\
    p(\truepm \given f, \bs{m}, s) &=
        f \, p_{\textrm{cl}}(\truepm \given \bs{m}, s)
        + (1-f) \, p_{\textrm{XD}}(\truepm)\\
    p_{\textrm{cl}}(\truepm \given \bs{m}, s) &=
        \norm(\truepm \given \bs{m}, s^2 \, \mathbb{I})
\end{align}
where $\norm(\bs{x} \given \bs{m}, \mat{C})$ represents the multidimensional normal distribution with mean $\bs{m}$ and covariance matrix $\mat{C}$, and $\mathbb{I}$ is the identity matrix.
Because all distributions are Gaussian, the per-star parameters (the true proper motions, \truepm) can be analytically marginalized out so that the per-star likelihood can be expressed as $p(\bs{\mu} \given f, \bs{m}, s, \mat{C}_\mu)$.
We assume that the measurements for each star, $n$, are independent so that the full likelihood of all $N$ stars given the parameters $(f, \bs{m}, s)$ is
\begin{align}
    p(\{\bs{\mu}_n\}_N \given \given f, \bs{m}, s, \{\mat{C}_{\mu, n}\}_N) &=
        \prod_n^N p(\bs{\mu}_n \given f, \bs{m}, s, \mat{C}_{\mu, n}) \quad .
        \label{eq:likelihood}
\end{align}

We use an ensemble Markov chain Monte Carlo (MCMC) sampler \citep[\texttt{emcee}][]{Goodman:XX, Foreman-Mackey:2013} to generate posterior samples over the parameters $(f, \bs{m}, s)$ using the likelihood defined above (Equation~\ref{eq:likelihood}), and assuming the following prior probability distributions: uniform over the domain $(-5, 5)~\masyr$ for each component of $\bs{m}$, uniform in $f$, and uniform in log-$s$ over the domain $-6 < \ln s < 4$ (with $s$ in units of \masyr).
We run the sampler with 32 walkers for 256 steps as burn-in, then reset the sampler and run for an additional 512 steps.
We then thin the resulting chains by taking every 16th sample, leaving a total of 1024 samples; We use these samples to estimate the median posterior parameter values and uncertainties, and to compute, for each star, posterior probabilities of belonging to the cluster mixture component.
For the young cluster, we find $\bs{m} = (-0.55,  0.44) \pm (0.03, 0.02)~\masyr$, $\ln s = -3.5 \pm 1$, and $f = 0.24 \pm 0.02$.
The bottom-left panel of the middle panels of \figurename~\ref{fig:pm-members} shows, in black (small black ellipse), the inferred median proper motion and proper-motion dispersion of the young cluster.
The proper-motion dispersion is consistent with zero, but is unmeasured; 95\% of the posterior samples have $s < 0.09~\masyr$.

The bottom-right panel of the middle panels of \figurename~\ref{fig:pm-members} shows all stars in the cluster field, but each marker is colored by the probability of (kinematically) belonging to the young cluster.
The right (large) panel of \figurename~\ref{fig:pm-members} shows the sky positions of all stars with $\textrm{probability} > 0.5$ of belonging to the cluster: The comoving, blue stars are also a clear spatial over-density in this region.
\figurename~\ref{fig:pm-members-cmd} shows \gaia\ color-magnitude diagrams for

% Notebook: Proper-motion-membership
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/pm-model.pdf}
\caption{\todo{APW}
}
\label{fig:pm-members}
\end{figure}

% Notebook: Proper-motion-membership
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/pm-members-cmd.pdf}
\caption{\todo{APW}
}
\label{fig:pm-members-cmd}
\end{figure}

@@

Fit isochrone? Or ask Dan Weisz?

@@

Query to retrieve \gaia\ data around the blue, comoving group found and discussed in \sectionname~\ref{sec:data}:
\begin{verbatim}
SELECT *
FROM gaiadr2.gaia_source
WHERE parallax < 1
    AND ra > (179-6) AND ra < (179+6)
    AND dec > (-29-6) AND dec < (-29+6)
\end{verbatim}

\software{
    \package{Astropy} \citep{astropy},
    \package{dustmaps}\footnote{\url{https://github.com/gregreen/dustmaps}},
    \package{emcee} \citep{emcee},
    \package{gala} \citep{gala},
    \package{IPython} \citep{ipython},
    \package{matplotlib} \citep{mpl},
    \package{numpy} \citep{numpy},
    \package{scipy} \citep{scipy}
}
